\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-02-05 paper.qmd
Add hexagonal grid starting point section

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltScurve
##| out-width: 100%

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
##| out-width: 100%

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons? The MSE plot have a steep slope at the beginning, indicating that a smaller benchmark causes a small amount of error. Then, the slope gradually increases or level up, indicating that a higher number of benchmark values generates a higher error. Using the reverse elbow method, when the benchmark value is set to 0.2, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that higher benchmark values remove the necessary bins as well which lead to distruct the 2D structure."
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelScurvermlgimp
#| fig-cap: "What does the 2D model, constructed using UMAP (applied to the S-curve dataset), look like in high dimensions with long edges? (<https://youtu.be/pSL_G_n7-fM>) (a) 2D model without removing the long edges, and (b) 2D model with long edges coloured by red. The default benchmark value is used to identify long edges."
##| out-width: 100%

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the long edges. What is the effective benchmark value to remove the long edges? $1.347$ is the chosen benchmark value to remove long edges. The value is selected through a visual inspection of the triangular mesh is conducted to verify whether it successfully eliminates long edges."
#| label: fig-diagnosticpltScurvelgrm
##| out-width: 100%

@@

#### Hexagonal grid starting point

Mainly hexagonal grid starts at the lowest x and y values assign for the grid. Another possibility is to shift the grid. That's another interesting parameter to investigate.

@@

ggplot(data = hex_full_count_df, aes(x = c_x, y = c_y)) +
  geom_point(color = "black") +
  geom_point(data = shifted_hex_coord_df, aes(x = c_x, y = c_y), color = "red")

ggplot(data = shifted_hex_coord_df, aes(x = x, y = y)) +
  geom_polygon(fill = NA, color = "#feb24c", aes(group = polygon_id)) +
  geom_polygon(data = hex_full_count_df, aes(x = x, y = y, group = polygon_id),
               fill = NA, color = "#bdbdbd")

## Before shift
ggplot(data = hex_full_count_df, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(aes(x = c_x, y = c_y, label = hexID)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff")

## After shift
ggplot(data = shifted_hex_coord_df, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(aes(x = c_x, y = c_y, label = hexID)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff")
```
Compute aic, mse after shifting and compare before and after shifting errors
And investigate how aic, mse change according to the shifted amount

>>>>

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltScurve
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
#| fig-pos: H
##| out-width: 100%

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons? The MSE plot have a steep slope at the beginning, indicating that a smaller benchmark causes a small amount of error. Then, the slope gradually increases or level up, indicating that a higher number of benchmark values generates a higher error. Using the reverse elbow method, when the benchmark value is set to 0.2, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that higher benchmark values remove the necessary bins as well which lead to distruct the 2D structure."
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelScurvermlgimp
#| fig-cap: "What does the 2D model, constructed using UMAP (applied to the S-curve dataset), look like in high dimensions with long edges? (<https://youtu.be/pSL_G_n7-fM>) (a) 2D model without removing the long edges, and (b) 2D model with long edges coloured by red. The default benchmark value is used to identify long edges."
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the long edges. What is the effective benchmark value to remove the long edges? $1.347$ is the chosen benchmark value to remove long edges. The value is selected through a visual inspection of the triangular mesh is conducted to verify whether it successfully eliminates long edges."
#| label: fig-diagnosticpltScurvelgrm
##| out-width: 100%
#| fig-pos: H

@@

#### Hexagonal grid origin

According to @Dan2023, the hexagonal binning is done by tessellating the $xy$ plane over the set (range($x$), range($y$)) (see @fig-scurveshifthexgridsexp (b)). In that case, bin centroids are defined as shown in @fig-scurveshifthexgridsexp (a) with gray colour. Rather than sticking to the typical hexagonal grid, introducing a meaningful shift in both the x and y directions presents an opportunity for an improved 2D model. Therefore, investigating this shift in the hexagonal grid is an important parameter to consider.

As shown in @fig-scurveshifthexgrids, the shifting influences the distribution of points and number of non-empty bins, impacting the resulting 2D model. According to @fig-diagnosticpltScurvehexbins, the 2D model with a total of $144$ bins applied to S-curve UMAP data does not require any shifting because the lowest MSE occurs when no shift is introduced.

@@

bin_centroids_shift <- ggplot(data = hex_full_count_df, aes(x = c_x, y = c_y)) +
  geom_point(color = "#bdbdbd") +
  geom_point(data = shifted_hex_coord_df, aes(x = c_x, y = c_y), color = "#feb24c") +
  coord_cartesian(xlim = c(-5, 8), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

hex_grid_shift <- ggplot(data = shifted_hex_coord_df, aes(x = x, y = y)) +
  geom_polygon(fill = NA, color = "#feb24c", aes(group = polygon_id)) +
  geom_polygon(data = hex_full_count_df, aes(x = x, y = y, group = polygon_id),
               fill = NA, color = "#bdbdbd") +
  coord_cartesian(xlim = c(-5, 8), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

## Before shift
before_shift_plot <- ggplot(data = hex_full_count_df, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(aes(x = c_x, y = c_y, label = hexID), size = 2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_void() +
  theme(legend.position="bottom", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

## After shift
after_shift_plot <- ggplot(data = shifted_hex_coord_df, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(aes(x = c_x, y = c_y, label = hexID), size = 2) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_equal() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) Visualization of bin centroids and (b) hexagonal grid, showcasing the configuration before (colored in gray) and after (colored in light orange) shifting the starting point by an identical amount applied in both x and y directions, with a shift value of $0.537285$."
#| label: fig-scurveshifthexgridsexp
##| out-width: 100%
#| fig-pos: H

bin_centroids_shift + hex_grid_shift +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

```{r}
#| echo: false
#| fig-cap: "Hexbin plots shows the distribution of points before and after shifting the starting point of the hexagonal grid generated for UMAP applied to the training S-curve dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after shifting (identical shift applied in both x and y directions, with a shift value of $0.537285$)."
#| label: fig-scurveshifthexgrids
##| out-width: 100%
#| fig-pos: H

before_shift_plot + after_shift_plot +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

#MSE_df <- dplyr::bind_rows(MSE_df_1, MSE_df_2, MSE_df_3, MSE_df_4, MSE_df_5)

#MSE_df$method <- factor(MSE_df$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

MSE_df <- read_rds("data/s_curve/s_curve_umap_shift_summary.rds") 
MSE_df$method <- factor(MSE_df$method, levels = c("UMAP"))

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0, linetype="dashed",
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("shift value")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 0, linetype="dashed",
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("shift value")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different shift values. In this illustration, an identical shift is applied in both the x and y directions. The Mean Squared Error (MSE) plot exhibits a global minimum at $0$, indicating that no additional shifting is necessary for the current hexagonal grid to yield a well-fitted model."
#| label: fig-diagnosticpltScurvehexbins
##| out-width: 100%
#| fig-pos: H

### 2024-02-05 paper.tex
Add hexagonal grid starting point section

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{hexagonal-grid-origin}{%
\subsubsection{Hexagonal grid origin}\label{hexagonal-grid-origin}}

According to \citet{Dan2023}, the hexagonal binning is done by
tessellating the \(xy\) plane over the set (range(\(x\)), range(\(y\)))
(see Figure~\ref{fig-scurveshifthexgridsexp} (b)). In that case, bin
centroids are defined as shown in
Figure~\ref{fig-scurveshifthexgridsexp} (a) with gray colour. Rather
than sticking to the typical hexagonal grid, introducing a meaningful
shift in both the x and y directions presents an opportunity for an
improved 2D model. Therefore, investigating this shift in the hexagonal
grid is an important parameter to consider.

As shown in Figure~\ref{fig-scurveshifthexgrids}, the shifting
influences the distribution of points and number of non-empty bins,
impacting the resulting 2D model. According to
Figure~\ref{fig-diagnosticpltScurvehexbins}, the 2D model with a total
of \(144\) bins applied to S-curve UMAP data does not require any
shifting because the lowest MSE occurs when no shift is introduced.

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

\caption{\label{fig-scurveshifthexgridsexp}(a) Visualization of bin
centroids and (b) hexagonal grid, showcasing the configuration before
(colored in gray) and after (colored in light orange) shifting the
starting point by an identical amount applied in both x and y
directions, with a shift value of \(0.537285\).}

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

\caption{\label{fig-scurveshifthexgrids}Hexbin plots shows the
distribution of points before and after shifting the starting point of
the hexagonal grid generated for UMAP applied to the training S-curve
dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after
shifting (identical shift applied in both x and y directions, with a
shift value of \(0.537285\)).}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvehexbins-1.pdf}

\caption{\label{fig-diagnosticpltScurvehexbins}Goodness of fit
statistics from UMAP applied to training S-curve dataset with different
shift values. In this illustration, an identical shift is applied in
both the x and y directions. The Mean Squared Error (MSE) plot exhibits
a global minimum at \(0\), indicating that no additional shifting is
necessary for the current hexagonal grid to yield a well-fitted model.}
